---
# title: 'STAT 2670: Elementary Statistics with R'
# subtitle: "Fall 2023"
author: "Jieun Park"
# date: "2023-09-30"
# header-includes:
# - \usepackage{pdflscape}
# - \newcommand{\blandscape}{\begin{landscape}}
# - \newcommand{\elandscape}{\end{landscape}}
# output:
#   html_document:
#     fig_caption: yes
#     number_sections: yes
#   word_document: default
#   pdf_document:
#     fig_caption: yes
#     number_sections: yes
#     toc: true
#     toc_depth: 2
# urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```
## Hypothesis Testing


### Basic of Hypothesis Testing
We will use the following functions to perform hypothesis tests.
```{r, message=FALSE}
library(BSDA)
# prop.test(x, n, p = NULL,
#           alternative = c("two.sided", "less", "greater"),
#           conf.level = 0.95, correct = TRUE)

# t.test(x, y = NULL,
#       alternative = c("two.sided", "less", "greater"),
#       mu = 0, paired = FALSE, var.equal = FALSE,
#       conf.level = 0.95, ...)

# z.test(
#   x, y = NULL,
#   alternative = "two.sided",
#   mu = 0, sigma.x = NULL, sigma.y = NULL,
#   conf.level = 0.95)

```


We use `qnorm()` and `qt()` functions to calculate critical values. For example, we can obtain $z_{0.05}$ using the `qnorm(0.95)` for a normal distribution, and the critical value $t_{0.05, 5}$ using `qt(0.95, 5)` for a t-distribution with 5 degree of freedom with $\alpha=0.05$ as below. 
```{r}
qnorm(0.95)
qt(0.95, 5)
```



### Testing a Claim About a Proportion

`mtcars` dataset has data for 32 automobiles in 1973-1974 with 11 variables. Among these variable, we are interested to check if the proportion of V-shaped engine (`vs = 0`) is 0.5. That is, $H_0: p = 0.5$. We set the null hypothesis as follows: the population proportion of cars with a V-shaped engine (vs = 0) among all automobiles in 1973-1974 is equal to 0.5.
We first check if we can use a normal approximation to perform a proportion test. With a sample size of $n=32$ and a proportion of interest $p=0.5$, both the expected number of successes and failures are  $np= n(1-p) = 32\cdot 0.5 = 16$. Since they are greater than 5, we can apply the proportion test using a normal approximation. In our sample, the number of success (`vs=0`) is 18 and the sample proportion is 0.56.
```{r, message=FALSE}
data(mtcars)
attach(mtcars)
table(vs)
prop.table(table(vs))
```



We use one sample proportion test with `prop.test()` function if $np \ge 5$ and $n(1-p) \ge 5$  where $p$ is the null hypothesized proportion and $n$ is the sample size. The syntax is below if we want to test with a sample vector (categorical variable with two levels) for $H_0: p = p_0$ with $\alpha = 0.05$. `x` is the number of success, `n`is the sample size, and `p_0` is the null hypothesized proportion.
```{r}
# prop.test(x, n, p = p_0, conf.level=0.95, alternative=c("two.sided", "less", "greater"))

```
Depending on the alternative hypothesis $H_1$, we can choose one among `two.sided`, `less`, and `greater`.
Under $H_0: p = p_0$, we can use each `alternative` option for `prop.test()` function.\

1. $H_1: p = p_0$: `alternative = "two.sided"`\
2. $H_1: p = p_0$ :`alternative = "less"`\
3. $H_1: p = p_0$: `alternative = "less"`\

For the proportion of `vs` , we test for the proportion of `vs = 0` with $H_0: p = 0.5$ and $\alpha = 0.05$.

#### Two-sided Proportion Test
$$H_0: p = 0.5 \quad \textrm{ vs }\quad H_1: p \ne 0.5$$
```{r}
res <- prop.test(x=18, n=32, p = 0.50, alternative = "two.sided", conf.level = 0.95)
res
```
**Decision:**

* **P-Value**: we fail to reject the null hypothesis since p-value `r round(res$p.value, 3)` is greater than $\alpha=0.05$.\
* **Critical Value**: the z-test statistic $z=$ `r round(sqrt(res$statistic), 3)` is closer to 0 than the critical values. Thus, we fail to reject the null hypothesis. \
```{r}
# the critical value can be calculated by the following code.
c(qnorm(0.025), qnorm(0.975))
```
* **Confidence Interval**: the claimed proportion 0.5 falls within the confidence interval of (`r round(res$conf.int[1],3)`, `r round(res$conf.int[2],3)`). Thus we fail to reject the null hypothesis.\


#### One-sided Proportion Test
$$H_0: p = 0.5 \quad \textrm{ vs }\quad H_1: p > 0.5$$
```{r}
res <- prop.test(x=18, n=32, p = 0.50, alternative = "greater", conf.level = 0.95)
res
```
**Decision:**

* **P-Value**: we fail to reject the null hypothesis since p-value `r round(res$p.value, 3)` is greater than $\alpha=0.05$.\
* **Critical Value**: the test statistic $z=$ `r round(sqrt(res$statistic), 3)` does not fall in the critical region which is greater than $z_{0.05}$ = `r round(qnorm(0.95),3)`. Thus, we fail to reject the null hypothesis. \
```{r}
# the critical value can be calculated by the following code.
qnorm(0.95)
```
* **Confidence Interval**: the claimed proportion 0.5 falls within the confidence interval of (`r round(res$conf.int[1],3)`, `r round(res$conf.int[2],3)`). Thus we fail to reject the null hypothesis.\



### Tesing a Claim About a Mean

#### Unknown $\sigma$ with Normality Assumption

We use one sample t-test with `t.test()` function when we assume normality for population or the sample size is large enough. The syntax is below if we want to test with a sample vector (variable) `x` for $H_0: \mu = m$ with $\alpha = 0.05$. 
```{r}
# t.test(x, mu= m, conf.level=0.95, alternative=c("two.sided", "less", "greater"))

```
Depending on the alternative hypothesis $H_1$, we can choose one among `two.sided`, `less`, and `greater`.
Under $H_0: \mu = m$, use each `alternative` option for `t.test()` function.\

1. $H_1: \mu \ne m$: `alternative = "two.sided"`\
2. $H_1: \mu < m$ :`alternative = "less"`\
3. $H_1: \mu > m$: `alternative = "less"`\

As an example, we test for `mpg` with $H_0: \mu = 22$. That is, we test if the population mean of `mpg` is equal to 22. `mtcars` cars have 32 samples and we can understand that we have a large enough sample to use t-test with $\alpha = 0.05$.

##### Two-sided t-test
$$H_0: \mu = 22 \quad \textrm{ vs }\quad H_1: \mu \ne 22$$
```{r}
res <- t.test(mpg, mu=22, alternative = "two.sided", conf.level = 0.95)
res
```
**Decision:**

* **P-Value**: we fail to reject the null hypothesis since p-value `r round(res$p.value, 3)` is greater than $\alpha=0.05$.\
* **Critical Value**: the test statistic $t=$ `r round(res$statistic, 3)` is closer to 0 than the critical values. Thus, we fail to reject the null hypothesis. \
```{r}
# the critical value can be calculated by the following code.
c(qt(0.025, df=31), qt(0.975, df=31))
```
* **Confidence Interval**: the claimed mean 22 falls within the confidence interval of (`r round(res$conf.int[1],3)`, `r round(res$conf.int[2],3)`). Thus we fail to reject the null hypothesis.\

##### One-sided t-test $H_1: \mu < m$
$$H_0: \mu = 22 \quad \textrm{ vs }\quad H_1: \mu < 22$$
```{r}
res <- t.test(mpg, mu=22, alternative = "less", conf.level = 0.95)
res
```
**Decision:**

* **P-Value**: we reject the null hypothesis since p-value `r round(res$p.value, 3)` is less than $\alpha=0.05$.\
* **Critical Value**: the test statistic $t=$ `r round(res$statistic, 3)` falls in the critical region which is less than $t_{0.05, 31}$ = `r round(qt(0.05, df=31),3)`. Thus, we reject the null hypothesis. \
```{r}
# the critical value can be calculated by the following code.
qt(0.05, df=31)
```
* **Confidence Interval**: the claimed mean does not fall within the confidence interval of ($-\infty$, `r round(res$conf.int[2],3)`). Thus we reject the null hypothesis.\



#### Known $\sigma$ with Normality Assumption

We use one sample z-test or normal test with `z.test()` function when we assume normality for population with known population standard deviation $\sigma$. The syntax is below if we want to test with a sample vector (variable) `x` for $H_0: \mu = m$ with $\alpha = 0.05$ and known `sigma`. 
```{r}
#library(BSDA)
# z.test(x, mu = m, sigma.x = sigma, conf.level = 0.95, alternative = c("two.sided", "less", "greater"))

```
Depending on the alternative hypothesis $H_1$, we can choose one among `two.sided`, `less`, and `greater`.
Under $H_0: \mu = m$, use each `alternative` option for `t.test()` function.\

1. $H_1: \mu \ne m$: `alternative = "two.sided"`\
2. $H_1: \mu < m$ :`alternative = "less"`\
3. $H_1: \mu > m$: `alternative = "less"`\

For example, we test for `mpg` with $H_0: \mu = 22$. Assume `mpg` follows a normal distribution with $\sigma = 6$, then we can use z-test with $\alpha = 0.05$.

##### Two-sided z-test
$$H_0: \mu = 22 \quad \textrm{ vs }\quad H_1: \mu \ne 22$$
```{r}
library(BSDA)
res <- z.test(mpg, mu=22, sigma.x = 6, alternative = "two.sided", conf.level = 0.95)
res
```
**Decision:**

* **P-Value**: we fail to reject the null hypothesis since p-value `r round(res$p.value, 3)` is greater than $\alpha=0.05$.\
* **Critical Value**: the test statistic $z=$ `r round(res$statistic, 3)` is closer to 0 than the critical values. Thus, we fail to reject the null hypothesis. \
```{r}
# the critical value can be calculated by the following code.
c(qnorm(0.025), qnorm(0.975))
```
* **Confidence Interval**: the claimed mean 22 falls within the confidence interval of (`r round(res$conf.int[1],3)`, `r round(res$conf.int[2],3)`). Thus we fail to reject the null hypothesis.\

##### One-sided z-test $H_1: \mu < m$
$$H_0: \mu_{mpg} = 22 \quad \textrm{ vs }\quad H_1: \mu_{mpg} < 22$$
```{r}
res <- z.test(mpg, mu=22, sigma.x = 6, alternative = "less", conf.level = 0.95)
res
```
**Decision:**

* **P-Value**: we reject the null hypothesis since p-value `r round(res$p.value, 3)` is less than $\alpha=0.05$.\
* **Critical Value**: the test statistic $z=$ `r round(res$statistic, 3)` falls in the critical region which is less than $z_{0.05}$ = `r round(qnorm(0.05),3)`. Thus, we reject the null hypothesis. \
```{r}
# the critical value can be calculated by the following code.
qnorm(0.05)
```
* **Confidence Interval**: the claimed mean does not fall within the confidence interval of ($-\infty$, `r round(res$conf.int[2],3)`). Thus we reject the null hypothesis.\


