---
title: 'STAT 2670: Elementary Statistics with R'
subtitle: "Fall 2023"
author: "Jieun Park"
date: "2023-09-30"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
output:
  html_document:
    fig_caption: yes
    number_sections: yes
  word_document: default
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: true
    toc_depth: 2
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hypothesis Testing

### Basic of Hypothesis Testing

**Exercise:** A soft drink company claims that 75% of their customers prefer their new flavor over the old one. To test this claim, a random sample of 200 customers was taken, and 145 of them preferred the new flavor. Use $\alpha = 0.10$.

(a) Which function can you use to test this claim? \
  A. `t.test()`\
  B. `p.test()`\
  C. `z.test()`\
  D. `prop.test()`\
(b) What is the proper option for `alternative`?\
  A. `two.sided`\
  B. `less`\
  C. `greater`\
  D. None of the above\
(c) What is the proper value for `conf.level`?\
  A. 0.9\
  B. 0.1\
  C. 0.05\
  D. 0.95\

### Testing a Claim About a Proportion
**Exercise:** A soft drink company claims that 75% of their customers prefer their new flavor over the old one. To test this claim, a random sample of 200 customers was taken, and 145 of them preferred the new flavor. Use $\alpha = 0.05$.
Answer the questions based on the R results below.\
```{r, echo=FALSE}
prop.test(145, 200, p=0.75, alternative = 'two.sided', conf.level=0.95)

```


(a) Which code can you use to test this claim? \
  A. `prop.test(145, 200, p=0.75, alternative = 'two.sided', conf.level=0.95)`\
  B. `prop.test(200, 145, p=0.75, alternative = 'two.sided', conf.level=0.95)`\
  C. `prop.test(145, 200, p=75, alternative = 'two.sided', conf.level=0.05)`\
  D. `prop.test(200, 145, p=75, alternative = 'two.sided', conf.level=0.05)`\
(b) What is the p-value?\
  A. 0.54\
  B. 0.75\
  C. 0.46\
  D. 0.725\
(c) What is the point estimate of the proportion of the customers who prefer the new flavor over the old one?\
  A. 0.54\
  B. 0.75\
  C. 0.46\
  D. 0.725\
(d) What is the corresponding confidence interval?\
  A. (0.54, 0.75)\
  B. (0.46, 0.73)\
  C. (0.66, 0.78)\
  D. (0.54, 0.78)\
(e) What is the decision?\
  A. We reject the claim.\
  B. We accept the claim.\
  C. We cannot make a decision.\

### Tesing a Claim About a Mean
**Exercise:** A tutoring center claims that their new study program improves students' test scores greater than 85 in average significantly. To test this claim, a random sample of 20 students who completed the program was taken. Their test scores after completing the program are as follows: [85, 88, 92, 78, 91, 89, 87, 90, 93, 84, 88, 86, 92, 91, 94, 86, 89, 90, 85, 88]. We assume that the population follows a normal distribution. Use $\alpha = 0.05$.
Answer the questions based on the R results below assuming `x= c(85, 88, 92, 78, 91, 89, 87, 90, 93, 84, 88, 86, 92, 91, 94, 86, 89, 90, 85, 88)`\
```{r, echo=FALSE}
x= c(85, 88, 92, 78, 91, 89, 87, 90, 93, 84, 88, 86, 92, 91, 94, 86, 89, 90, 85, 88)
t.test(x, mu=85, alternative = 'greater', conf.level=0.95)

```
(a) Which code can you use to test this claim? \
  A. `t.test(x, mu=85, alternative = 'less', conf.level=0.95)`\
  B. `t.test(x, mu=85, alternative = 'two.sided', conf.level=0.95)`\
  C. `t.test(x, mu=85, alternative = 'greater', conf.level=0.05)`\
  D. `t.test(x, mu=85, alternative = 'greater', conf.level=0.95)`\
(b) What is the p-value?\
  A. 3.9727\
  B. 0.0004\
  C. 86.864\
  D. 88.3\
(c) What is the point estimate of the mean score from new study program?\
  A. 3.9727\
  B. 0.0004\
  C. 86.864\
  D. 88.3\
(d) What is the test statistic?\
  A. 3.9727\
  B. 0.0004\
  C. 86.864\
  D. 88.3\
(e) What is the critical value for the test?\
  A. `qt(0.95, 20)`\
  B. `qt(0.975, 20)`\
  C. `qt(0.95, 19)`\
  D. `qt(0.975, 19)`\
  
(e) What is the decision?\
  A. The new study program improves students' test scores greater than 85 in average. \
  B. We do not have enough evidence to support the new study program improves the scores greater than 85 in average. \
  C. We cannot make a decision.\

## Correlation and Regression

### Correlation
**Exercise:** Consider the following code. 
```{r}
x <- 1:20
y <- -(20:1)

```
How can we calculate the correlation coefficient between x and y?\
  A. `cov(x, y)`\
  B. `cor(x, y)`\
  C. `corr(x,y)`\
  D. None of the above\



### Regression
**Exercise:** Consider the following R result. 
```{r, echo=FALSE}
data(mtcars)
res <- lm(mpg ~ wt, mtcars)
summary(res)
```
(a) What is the response variable?\
  A. `mtcars`\
  B. `mpg`\
  C. `wt`\
  D. None of the above\
(b) What is the predictor variable?\
  A. `mtcars`\
  B. `mpg`\
  C. `wt`\
  D. None of the above\
(c) What is the slope estimate?\
  A. 37.2851\
  B. 1.8776\
  C. -5.3445\
  D. 3.046\
(d) What is the intercept estimate?\
  A. 37.2851\
  B. 1.8776\
  C. -5.3445\
  D. 3.046 \ 
(e) What is the predicted `mpg` when `wt` is 3?\
  A. 21.2516\
  B. 6.8727\
  C. 19.858\
  D. 91.38\




