---
# title: 'STAT 2670: Elementary Statistics with R'
# subtitle: "Chapter 6 and 7"
author: "Wen Tang"
# date: "2023-08-22"
# header-includes:
# - \usepackage{pdflscape}
# - \newcommand{\blandscape}{\begin{landscape}}
# - \newcommand{\elandscape}{\end{landscape}}
# output:
#   word_document: default
#   pdf_document:
#     fig_caption: yes
#     number_sections: yes
#     toc: true
#     toc_depth: 2
#   html_document:
#     fig_caption: yes
#     number_sections: yes
# urlcolor: blue
---

```{r set, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NORMAL PROBABILITY DISTRIBUTION
### THE standard normal distribution
#### Normal distribution graph (Displaying only).

```{r}
set.seed(123)                       # Set the seed for reproducibility
x <- rnorm(1000, mean = 0, sd = 1)  # Generate data for a standard normal distribution

# Plot the data with density curve
hist(x, prob = TRUE, col = "lightblue", main = "Standard Normal Distribution")
lines(density(x), col = "red", lwd = 2)
```

#### Find the probability (area) when z scores are given.

```{r}
# Find the area under the curve to the left of a certain value: P(z<1)
pnorm(1, mean = 0, sd = 1)
# Find the area under the curve to the right of a certain value: P(z>1)
1-pnorm(1, mean = 0, sd = 1)
# Find the area under the curve between two values: P(-1<z<1)
diff(pnorm(c(-1, 1), mean = 0, sd = 1))
```

#### Find z scores when the area is given.

```{r}
# Find the value with a certain area under the curve to its left: critical value 
alpha <- 0.05
qnorm(1-alpha, mean = 0, sd = 1) # find the critical Z score.
```

### REAL application of normal distribution
#### Convert an individual x value to a z-score

```{r}
x <- 80  # the individual value
mu <- 75  # the mean of the distribution 
sigma <- 10  # the standard deviation of the distribution 

# Calculate z-scores for the individual value using scale()
z_scores <- scale(x, center = mu, scale = sigma)
cat("Z-score:", z_scores, "\n") # print the z-score

z <- (x - mu) / sigma  # find the z-score by using the formula 
cat("Z =", z, "\n") # print the z-score
  
```
#### Find the probability when x value is given (page 269 Pulse Rates Question)

```{r}
x1 <- 60
x2 <- 80
mu <- 69.6
sigma <- 11.3
# Find the probability that X is less than 60: P(X<60)
pnorm(x1, mean = mu, sd = sigma)
# Find the probability that X is great than 80: P(X>80)
1-pnorm(x2, mean = mu, sd = sigma)
# Find the probability between two values: P(60<X<80)
diff(pnorm(c(x1, x2), mean = mu, sd = sigma))
```

#### Convert a z-score back to x value

```{r}
z <- 1.96  # the z-score
mu <- 100  # the mean of the distribution
sigma <- 15  # the standard deviation of the distribution
x <- z * sigma + mu  # convert the z score to individual x value using formula
cat("X =", x, "\n")  # print the individual x value
```

### SAMPLING distributions and estimators (Displaying only/Optional)
#### general behavior of sampling distribution of the sample proportion

```{r}
# Set the seed for reproducibility
set.seed (123)
# Generate data
n <- 10  # sample size
p <- 0.5  # population proportion
samples <- replicate(50000, rbinom(1, size = n, prob = p))

# Calculate sample proportions
sample_props <- samples / n

# Plot the histogram

hist(sample_props, breaks = seq( 0, 1, by = 0.1 ), col = "lightblue", main = "Sampling Distribution of Sample Proportion")
```

#### general behavior of sampling distribution of the sample mean

```{r}
#input the parameter values
mu <- 3.5    
sigma <- 1.7       
n <- 5         
# Simulate sampling distribution
sample_means <- replicate(10000, mean(rnorm(n, mu, sigma)))

# Create a histogram of the sampling distribution of the sample mean
hist(sample_means, breaks ="FD",  main = "Sampling Distribution of Sample Mean", xlab = "Sample Mean", ylab = "Frequency", col = "lightblue", border = "black")
```

#### general behavior of sampling distribution of the sample variance

```{r}
mu <- 4    # True population mean
sigma <- 8       # Population standard deviation
sample_size <- 10         # Sample size
num_samples <- 10000       # Number of samples
# Function to calculate sample variance
sample_variance <- function(sample) {
  n <- length(sample)
  mean_sample <- mean(sample)
  sum_squared_deviations <- sum((sample - mean_sample)^2)
  return(sum_squared_deviations / (n - 1))
}
# Simulate sampling distribution
sample_variances <- replicate(num_samples, sample_variance(rnorm(sample_size, mu, sigma)))

# Create a histogram of the sampling distribution of sample variance
hist(sample_variances, breaks = "FD", freq = FALSE, main = "Sampling Distribution of Sample Variance",
     xlab = "Sample Variance", ylab = "Frequency", col = "lightblue", border = "black")
```

### THE central limit theorem
#### Find the probability when individual value is used (Page 292 Ejection Seat Question)

```{r}
mu <- 171 # population mean
sigma <- 46 # population standard deviation
n <- 25 # sample size
x_lower <- 140
x_upper <- 211

# Find the probability between two X values
probability_range <- diff(pnorm(c(x_lower, x_upper), mean = mu, sd = sigma))
probability_range
```

#### Find the probability when sample mean is used (Page 292 Ejection Seat Question)

```{r}
# Find the probability between two mean values $x/bar$ (CLT)
standard_error <- sigma / sqrt(n) # Calculate the standard error of the sample mean
probability_range <- diff(pnorm(c(x_lower, x_upper), mean = mu, sd = standard_error))# Find the probability  
probability_range
```

## ESTIMATING PARAMETERS AND DETERMINGING SAMPLE SIZES
### ESTIMATING a population proportion (Page 313 Online Course Example)
#### Getting the CI directly 
```{r}
p_hat <- 0.53 # 0.53 for 53% sample proportion
n <- 950 # sample size
success <- n*p_hat # number of success

# Calculate a 95% confidence interval for the population proportion
result <- prop.test(success, n, conf.level = 0.95)

# Extract the confidence interval
conf_interval <- result$conf.int
# Print the confidence interval
cat("Confidence Interval:", conf_interval[1], "to", conf_interval[2], "\n")
```

#### Getting the CI step by step
1.Critical value

```{r}
# Confidence level (e.g., 0.95 for 95% confidence)
confidence_level <- 0.95  
# get alpha value
alpha <- 1-confidence_level

# Find the critical Z-value using qnorm()
critical_z <- qnorm (1 - alpha/2)
# Print the result
cat("Critical Z =", critical_z, "\n")
```

2. Margin of error

```{r}

# Calculate the standard error
standard_error <- sqrt((p_hat * (1 - p_hat)) / n)
# Calculate the margin of error
margin_of_error <- critical_z * standard_error
# Print the result
cat("E=", margin_of_error, "\n")
```

3. Confidence interval

```{r}
# Calculate the confidence interval
confidence_interval <- c (p_hat - margin_of_error,
                          p_hat + margin_of_error)

# Print the confidence interval
cat("Confidence Interval:", confidence_interval[1], "to", confidence_interval[2], "\n")
```

### ESTIMATING a population mean 
#### Get the CI directly with Original data values are given. (Page 343 Mercury question)

```{r}

# Calculate a 98% confidence interval for the population mean
#Sample data 
mercury <- c(0.56, 0.75, 0.10, 0.95, 1.25, 0.54, 0.88)
result <- t.test(mercury,conf.level = 0.98)

# Extract the confidence interval
conf_interval <- result$conf.int

# Print the confidence interval
cat("Confidence Interval:", conf_interval[1], "to", conf_interval[2], "\n")
```

#### Get the CI step by step with given mean and standard deviation (Page 341 Hershey kisses question)

1. Critical value

```{r}
confidence_level <- 0.99  # Confidence level (e.g., 0.99 for 99% confidence)
alpha <- 1- confidence_level
n <- 32         # Sample size

# Calculate the degrees of freedom
degrees_of_freedom <- n - 1

# Find the critical t-value using qt()
critical_t <- qt(1 - alpha/ 2, df = degrees_of_freedom)

# Print the result
cat("Critical t-value for degrees of freedom =", degrees_of_freedom, "and confidence level =", confidence_level, ":", critical_t, "\n")
```

2. Margin of error

```{r}
# Given sample standard deviation (this is s value)
sample_standard_deviation <- 0.1077

# Calculate the standard error
standard_error <- sample_standard_deviation / sqrt(n)

# Calculate the margin of error
margin_of_error <- critical_t * standard_error

# Print the result
cat("Margin of Error for confidence level =", confidence_level, "and sample size =", n, ":", margin_of_error, "\n")
```

3. Confidence interval

```{r}
x_bar<- 4.5210       # Sample mean

# Calculate the lower and upper bounds of the confidence interval
lower_bound <- x_bar - margin_of_error
upper_bound <- x_bar + margin_of_error

# Print the result
cat("Confidence Interval:", lower_bound, "to", upper_bound, "\n")
```

### ESTIMATING a population variance (body temperature example page 353)
#### Critical values

```{r}
confidence_level <- 0.95  # Confidence level ( 0.95 for 95% confidence)
alpha <- 1- confidence_level
sample_size <- 106          # Sample size
degrees_of_freedom <- sample_size - 1  # Degrees of freedom for the chi-squared distribution

# Find the critical values using the chi-squared distribution
lower_critical_value <- qchisq(1-alpha/2, df = degrees_of_freedom)
upper_critical_value <- qchisq(alpha/2, df = degrees_of_freedom)

# Print the results
cat("Lower Critical Value:", lower_critical_value, "\n")
cat("Upper Critical Value:", upper_critical_value, "\n")
```

#### Confidence interval 

```{r}

sample_standard_deviation <- 0.62                  # sample standard deviation s
sample_variance <- sample_standard_deviation^2     # Sample variance

# Calculate the confidence interval for variance
confidence_interval <- c(((sample_size - 1) * sample_variance) / lower_critical_value,
                         ((sample_size - 1) * sample_variance) / upper_critical_value)

# Print the confidence interval
confidence_interval
```

## SAMPLE QUESTIONS FOR CHAPTER 6 AND 7
### SECTION 6.1
#### Bone density scores are normally distributed with a mean of 0 and a standard deviation of 1. Find the probability of the given bone density test scores. Please use r instead of table and round your answers to four decimal places.
1. Less than -2.00
2. Greater than 2.33
3. Between -0.77 and 1.42

#### Bone density scores are normally distributed with a mean of 0 and a standard deviation of 1.Find the bone desity test scores corresponding to the given information. Round your answer to two decimal places.
1. Find the 99th percentile $P_{99}$. This is the bone density score separating the bottom 99% from the top 1%.
2. Find the bone density scores that are the three Quartiles: $Q_1$, $Q_2$,$Q_3$.

#### Find the indicated critical value. Round results to two decimal places.
1. $Z_{0.25}$
2. $Z_{0.02}$
3. $Z_{0.06}$

### SECTION 6.2
#### The IQ test scores of adults are normally distributed with a mean of 100 and a standard deviation of 15 (As on the Wechsler IQ test).
1. Find the probability that a person has IQ score greater than 125.
2. Find the probability that a person has IQ score between 90 and 105.
3. Find the $P_{90}$, which is the IQ score to separating the bottom 90% from the top 10%.

### SECTION 6.4
#### Assume that weights of men are normally distributed with a mean of 189 lb and a standard deviation of 39 lb.
1. If one man is randomly selected, What is the probability that his weight exceeds 140 lb.
2. If 30 men are randomly selected, what is the probability that their mean weight exceeds 140 lb.


### SECTION 7.1
#### One of Mendel's famous genetics experiments yielded 580 peas, with 428 of them green and 152 yellow.
Find a 99% confidence interval estimate of the percentage of green peas.
 a) Find the critical value.
 b) Find the margin of error.
 c) Find the confidence interval

### SECTION 7.2
#### The summary statistics for the weights of Pepsi in randomly selected cans are n=36, $\bar{x}$=0.82410 lb, s=0.00570 lb. Use a confidence level of 95%
 a) Find the critical value.
 b) Find the margin of error.
 c) Find the confidence interval 

### SECTION 7.3
#### Assume the weights of dollar coins are normally distributed. Find the 95% confidence interval given n=20, s= 0.04111.
 a) Find the critical values
 b) Find the confidence interval.
